An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-07-06 main.tex
intro

The story gets more complicated with $\AcaOrBc{}$, where we have a choice of two
paths, up or down, and where that choice affects the possibility of taking
either of the two loops. In that case, we would have the image $2l_a + l_c + l_c' + l_b + l_b'$.
If we apply standard linear reasoning on the equations (e.g. $1 + {l_b}' = l_a +
l_b$, or $l_a + l_c' = 1$), we can reduce the number of bound
($\exists$-quantified) variables somewhat, obtaining the relations in
\cref{fig:example-simplify-1}, and the corresponding over-approximation on
the $f$-image $2 + l_c + 2l_b'$. It is an over-approximation because it only
considers the number of times each transition is taken and the existence of a
path from an initial to an accepting state, but omits the connectivity of that
path in the presence of loops. For example, no valid path would have both $l_c >
0$ and $l_b' > 0$, but no such restriction exists in the formula above.

@@

Still, this approximate solution allows us to perform further reasoning. Since
we know that the length of a string accepted by the intersection of the two
automata's languages must be the same, we also know that their images (string
lengths) must be the same under $f$, e.g. that $r_\Sigma + 1 + r_\Sigma' = 2 +
l_c + 2l_b' \iff r_\Sigma + r_\Sigma' = 1 + l_c + 2l_b'$. This in turn allows us
to replace one additional variable, e.g. $r_\Sigma = 1 + l_c + 2l_b' -
r_\Sigma'$ (see \cref{fig:example-length-reduced}). Note that the two automata
are now in contact, allowing us to propagate reasoning on one to the other.

@@

This is as far as we can get with this reasoning. We must now choose: either we
compute (materialise) the product and continue our reasoning, or we perform a
case split. Let us try a case split in order to put off computing a potentially
large product. 1 an early transition in the largest automaton with loops,
in our case $\FromLabelTo{S}{a}{A}$ of $\AcaOrBc$. This creates two
cases: $1-l_c' = 0$ and $1-l_c' > 0$. We start with the first case, and after
propagating the now known values and removing any transition that becomes zero,
we get the automaton of \cref{fig:aca-or-bc-length-split-1}.

@@

  We immediately notice that the self-loop of state $A$ is now disconnected and
  must be removed. This condition can be detected efficiently using standard
  forwards/backwards reachability from the initial and accepting states
  respectively. In other words, $2k+1 = 0$ for this case and is implied by the
  splitting constraint. However, no choice of integer $k$  will satisfy this
  equation. Therefore, we close this branch and proceed with the other one --
  where $1-l_c' > 0$. Note that we do so without ever computing a potentially
  expensive product of automata.

@@

Plugging in the assumption $1-l_c' > 0$ (and its consequence that $l_c' = 0$),
we also get a reduced variant of $\AcaOrBc$, as seen in
\cref{fig:aca-or-bc-length-split-2}. Note that there is now no way to select any
transition variable so as to disconnect a loop from a path between $S$ and $F$,
and we can therefore stop worrying about enforcing connectivity.

\begin{figure}[t]
  \centering 
    \includegraphics[width=0.5\linewidth]{length_split_product}
    \caption{The product of $\AcaOrBc$ and $\SomethingCSomething$ after
    case splitting and removal of dead states.}\label{fig:length-split-product}
  \end{figure}

Before continuing, we need to tie the inequalities for the transitions of the
automata of the product to the new transition counting variables we introduced
in the product automaton to ensure that they are consistent. In our case, we obtain the following equations from the principle that a transition variable in either automaton must be equal to the sum of variables in the resulting edges of the product, which gives the following two interesting inequality from $\AcaOrBc{}$'s transition $\FromLabelTo{sawA}{c}{sawA}$: $2k + 1 = r_1 + r_2 + 1$

Note how this conserves the requirement that the number of loop iterations be
odd even when the same original loop appears three times in the product, twice
as a self-loop and once as a regular state transition. Using the same principle
as before for computing the length by adding up the transition counters, we
obtain the quantified formula $1 + r_1 + 1 = r_2 + 1 = r_1 + r_2 + 3 = 2k + r_1
+ 3$ for the length. Eliminating the final quantifier gives us the image of
$f(\AcaOrBc \times \SomethingCSomething)$ modulo odd lengths: $3 + 2k$, for some
integer $k$.

@@

  \begin{minipage}[b]{0.45\linewidth}
    \begin{equation*}
      \begin{aligned}
        & \Filter(\FromLabelTo{start}{a}{sawA}) & = \TransitionVar_1 \\
        & \Filter(\FromLabelTo{start}{b}{sawB}) & = \TransitionVar_2 \\
        & \Filter(\FromLabelTo{sawA}{c}{sawA})  & = \TransitionVar_3  \\
        & \Filter(\FromLabelTo{sawB}{b}{start}) & = \TransitionVar_4 \\
        & \Filter(\FromLabelTo{sawA}{a}{final}) & = \TransitionVar_5 \\
        & \Filter(\FromLabelTo{sawB}{c}{final}) & = \TransitionVar_6 \\
      \end{aligned}
    \end{equation*}    
  \end{minipage}
  \hspace{0.5cm}
  \begin{minipage}[b]{0.45\linewidth}
    \begin{equation*}
      \begin{aligned}
        % start
        &  \TransitionVar_4 = \TransitionVar_5 + \TransitionVar_2 - 1 \\
        % sawA
        & \TransitionVar_1 = \TransitionVar_5 \\
        % sawB
        & \TransitionVar_2 = \TransitionVar_4 + \TransitionVar_6 \\
        % final
        & \TransitionVar_5 + \TransitionVar_6 = \FinalStateVar_1 \\
        & \FinalStateVar_1 = 1 \\
      \end{aligned}  
    \end{equation*}    
  \end{minipage}
  \caption{Equivalences defining $\Filter$ and $\FlowEq(\AcaOrBc{}, \Filter)$
  respectively.}\label{fig:example:single:equivalences}
  \end{figure}

Propagating the equivalences from \FlowEq{} gives us (under the implicit
assumption that every RHS is $\geq 0$):
\begin{equation*}
  \begin{aligned}
    & \Filter(\FromLabelTo{start}{a}{sawA}) & = 1 - \TransitionVar_6 \\
    & \Filter(\FromLabelTo{start}{b}{sawB}) & = \TransitionVar_4 + \TransitionVar_6 \\
    & \Filter(\FromLabelTo{sawA}{c}{sawA})  & = \TransitionVar_3  \\
    & \Filter(\FromLabelTo{sawB}{b}{start}) & = 2\TransitionVar_6 + \TransitionVar_4 \\
    & \Filter(\FromLabelTo{sawA}{a}{final}) & = 1 - \TransitionVar_6 \\
    & \Filter(\FromLabelTo{sawB}{c}{final}) & = \TransitionVar_6 \\
  \end{aligned}
\end{equation*}

@@

We return again to our example in \cref{sec:introduction:parikh}, where we
compute the whole Parikh image of the product of $\AcaOrBc{}$ and
$\SomethingCSomething$ under the constraint that there are more instances of
letters a than b. We omit the definition of the map $\Map$ that maps each
transition to the corresponding increment vector (e.g.
\Map($\FromLabelTo{start}{a}{sawA}) = \begin{bmatrix}1\\0\\0\end{bmatrix}$),
since it should be apparent by now.

@@

% Propagated equations
    \begin{equation*}
      \begin{aligned}
        & \Filter(\FromLabelTo{start}{a}{sawA}) & = 1 - \TransitionVar_6 \\
        & \Filter(\FromLabelTo{start}{b}{sawB}) & = \TransitionVar_4 + \TransitionVar_6 \\
        & \Filter(\FromLabelTo{sawA}{c}{sawA})  & = \TransitionVar_3  \\
        & \Filter(\FromLabelTo{sawB}{b}{start}) & = 2\TransitionVar_6 + \TransitionVar_4 \\
        & \Filter(\FromLabelTo{sawA}{a}{final}) & = 1 - \TransitionVar_6 \\
        & \Filter(\FromLabelTo{sawB}{c}{final}) & = \TransitionVar_6 \\
        % Other automaton
        & \Filter(\FromLabelTo{start}{\Sigma}{start}) & = \TransitionVar_{7a}, \TransitionVar_{7b}, \TransitionVar_{7c} \\
        & \Filter(\FromLabelTo{start}{c}{final}) & = 1 \\
        & \Filter(\FromLabelTo{final}{\Sigma}{final}) & = \TransitionVar_{9a}, \TransitionVar_{9b}, \TransitionVar_{9c} \\
      \end{aligned}
    \end{equation*}

@@

  synchronised {
  automaton aca_or_bc {
    init start;

    // We use ASCII values here, this is for lowercase a
    start -> sawA [97] { l_a += 1 };
    start -> sawB [98] { l_b += 1 };

    sawB -> start  [98] { l_b += 1 };
    sawB -> final [99] { l_c += 1 };

    sawA -> sawA [99] { l_c += 1 };
    sawA -> final [97] { l_a += 1 };

@@

  automaton something_c_something {
      init start;

      // Special short-hand value for the whole Unicode alphabet.
      start -> start [any] ;
      start -> final [99] { r_c += 1 };

      final -> final [any];

      accepting final;
  };
  };

>>>>

The story gets more complicated with $\AcaOrBc{}$, where we have a choice of two
paths, up or down, and where that choice affects the possibility of taking
either of the two loops. In that case, we would have the image $2l_a + l_c + l_c' + l_b + l_b'$.
If we apply standard linear reasoning on the equations (e.g. $1 + {l_b}' = l_a +
l_b$, or $l_a + l_c' = 1$), we can reduce the number of bound
($\exists$-quantified) variables somewhat, obtaining the relations in
\cref{fig:example-simplify-1}, and the corresponding over-approximation on
the $f$-image $2 + l_c + 2l_b'$. It is an over-approximation because it only
considers the number of times each transition is taken and the existence of a
path from an initial to an accepting state, but omits the connectivity of that
path in the presence of loops. For example, no valid path would have both $l_c >
0$ and $1 - l_c' = 0$, but no such restriction exists in the formula above.

@@

Still, this approximate view allows us to perform further reasoning. Since we
know that the length of a string accepted by the intersection of the two
automata's languages must be the same, we also know that their images must be
the same under $f$, e.g. that $r_\Sigma + 1 + r_\Sigma' = 2 + l_c + 2l_b' \iff
r_\Sigma + r_\Sigma' = 1 + l_c + 2l_b'$. This in turn allows us to replace one
additional variable, e.g. $r_\Sigma = 1 + l_c + 2l_b' - r_\Sigma'$ (see
\cref{fig:example-length-reduced}). Note that the two automata are now in
contact, allowing us to propagate reasoning on one to the other.

@@

This is as far as we can get with this reasoning. We must now choose: either we
compute (materialise) the product and continue our reasoning, or we perform a
case split. Let us try a case split in order to put off computing a potentially
large product.

We apply the principle of splitting and select an early transition in the
largest automaton with loops, in our case $\FromLabelTo{S}{a}{A}$ of $\AcaOrBc$.
This creates two cases: $1-l_c' = 0$ and $1-l_c' > 0$. We start with the first
case, and after propagating the now known values and removing any transition
that becomes zero, we get the automaton of \cref{fig:aca-or-bc-length-split-1}.

@@

  We immediately notice that the self-loop of state $A$ is now disconnected from
  the initial state and must be removed. This condition can be detected
  efficiently using standard forwards/backwards reachability from the initial
  and accepting states respectively. In other words, $2k+1 = 0$ for this case
  and is implied by the splitting constraint. However, no choice of integer $k$
  will satisfy this equation. Therefore, we close this branch and proceed with
  the other one -- where $1-l_c' > 0$. Note that we do so without ever computing
  a potentially expensive product of automata.

@@

Plugging in the assumption $1-l_c' > 0$ (and its consequence that $l_c' = 0$),
we also get a reduced variant of $\AcaOrBc$, as seen in
\cref{fig:aca-or-bc-length-split-2}. Note that there is now no way to select any
transition variable to disconnect a loop from a path between $S$ and $F$,
and we can therefore stop worrying about enforcing connectivity at all.

\begin{figure}[t]
  \centering 
    \includegraphics[width=0.5\linewidth]{length_split_product}
    \caption{The product of $\AcaOrBc$ and $\SomethingCSomething$ after
    case splitting and removal of dead states.}\label{fig:length-split-product}
    \Description[$\AcaOrBc{}$ with the transition from $B$ to $F$ removed.]{$\AcaOrBc{}$ with the transition from $B$ to $F$ removed. There now remains only one path to an accepting state.}
  \end{figure}

Before continuing, we need to tie the inequalities for the transitions of the
automata of the product to the new transition counting variables we introduced
in the product automaton to ensure that they are consistent. In our case, we
obtain the following equations from the principle that a transition variable in
either automaton must be equal to the sum of variables in the resulting edges of
the product, which gives the following two interesting inequalities from
$\AcaOrBc{}$'s transition $\FromLabelTo{A}{c}{A}$: $2k + 1 = r_1 + r_2 + 1$

Note how this conserves the requirement that the number of loop iterations be
odd even when the same original loop appears three times in the product, twice
as a self-loop and once as a regular state transition. Using the same principle
as before for computing the length by adding up the transition counters, we
obtain the quantified formula $1 + 2k + 1 + 1 + r_b + r_b= 3 + 2k + 2r_b$ for
the length. Eliminating the final quantifier gives us the image of $f(\AcaOrBc
\times \SomethingCSomething)$ modulo odd lengths: $3 + 2k$, for some integer
$k$.

@@

  \begin{minipage}[b]{0.45\linewidth}
    \begin{equation*}
      \begin{aligned}
        & \Filter(\FromLabelTo{S}{a}{A}) & = \TransitionVar_1 \\
        & \Filter(\FromLabelTo{S}{b}{B}) & = \TransitionVar_2 \\
        & \Filter(\FromLabelTo{A}{c}{A})  & = \TransitionVar_3  \\
        & \Filter(\FromLabelTo{B}{b}{S}) & = \TransitionVar_4 \\
        & \Filter(\FromLabelTo{A}{a}{F}) & = \TransitionVar_5 \\
        & \Filter(\FromLabelTo{B}{c}{F}) & = \TransitionVar_6 \\
      \end{aligned}
    \end{equation*}    
  \end{minipage}
  \hspace{0.5cm}
  \begin{minipage}[b]{0.45\linewidth}
    \begin{equation*}
      \begin{aligned}
        % S
        &  \TransitionVar_4 = \TransitionVar_5 + \TransitionVar_2 - 1 \\
        % A
        & \TransitionVar_1 = \TransitionVar_5 \\
        % B
        & \TransitionVar_2 = \TransitionVar_4 + \TransitionVar_6 \\
        % F
        & \TransitionVar_5 + \TransitionVar_6 = \FinalStateVar_1 \\
        & \FinalStateVar_1 = 1 \\
      \end{aligned}  
    \end{equation*}    
  \end{minipage}
  \caption{Equivalences defining $\Filter$ and $\FlowEq(\AcaOrBc{}, \Filter)$
  respectively.}\label{fig:example:single:equivalences}
  \end{figure}

Propagating the equivalences from \FlowEq{} gives us (under the implicit
assumption that every RHS is $\geq 0$):
\begin{equation*}
  \begin{aligned}
    & \Filter(\FromLabelTo{S}{a}{A}) & = 1 - \TransitionVar_6 \\
    & \Filter(\FromLabelTo{S}{b}{B}) & = \TransitionVar_4 + \TransitionVar_6 \\
    & \Filter(\FromLabelTo{A}{c}{A})  & = \TransitionVar_3  \\
    & \Filter(\FromLabelTo{B}{b}{S}) & = 2\TransitionVar_6 + \TransitionVar_4 \\
    & \Filter(\FromLabelTo{A}{a}{F}) & = 1 - \TransitionVar_6 \\
    & \Filter(\FromLabelTo{B}{c}{F}) & = \TransitionVar_6 \\
  \end{aligned}
\end{equation*}

@@

We return again to our example in \cref{sec:introduction:parikh}, where we
compute the whole Parikh image of the product of $\AcaOrBc{}$ and
$\SomethingCSomething$ under the constraint that there are more instances of
letters a than b. We omit the definition of the map $\Map$ that maps each
transition to the corresponding increment vector (e.g.
\Map($\FromLabelTo{S}{a}{A}) = \begin{bmatrix}1\\0\\0\end{bmatrix}$),
since it should be apparent by now.

@@

% Propagated equations
    \begin{equation*}
      \begin{aligned}
        & \Filter(\FromLabelTo{S}{a}{A}) & = 1 - \TransitionVar_6 \\
        & \Filter(\FromLabelTo{S}{b}{B}) & = \TransitionVar_4 + \TransitionVar_6 \\
        & \Filter(\FromLabelTo{A}{c}{A})  & = \TransitionVar_3  \\
        & \Filter(\FromLabelTo{B}{b}{S}) & = 2\TransitionVar_6 + \TransitionVar_4 \\
        & \Filter(\FromLabelTo{A}{a}{F}) & = 1 - \TransitionVar_6 \\
        & \Filter(\FromLabelTo{B}{c}{F}) & = \TransitionVar_6 \\
        % Other automaton
        & \Filter(\FromLabelTo{S}{\Sigma}{S}) & = \TransitionVar_{7a}, \TransitionVar_{7b}, \TransitionVar_{7c} \\
        & \Filter(\FromLabelTo{S}{c}{F}) & = 1 \\
        & \Filter(\FromLabelTo{F}{\Sigma}{F}) & = \TransitionVar_{9a}, \TransitionVar_{9b}, \TransitionVar_{9c} \\
      \end{aligned}
    \end{equation*}

@@

  synchronised {
  automaton aca_or_bc {
    init S;

    // We use ASCII values here, this is for lowercase a
    S -> A [97] { l_a += 1 };
    S -> B [98] { l_b += 1 };

    B -> S  [98] { l_b += 1 };
    B -> final [99] { l_c += 1 };

    A -> A [99] { l_c += 1 };
    A -> final [97] { l_a += 1 };

@@

  automaton something_c_something {
      init S;

      // Special short-hand value for the whole Unicode alphabet.
      S -> S [any] ;
      S -> F [99] { r_c += 1 };

      F -> F [any];

      accepting F;
  };
  };
